Geographical Python Teaching Resources: GeoPyTeR

### 2020-02-24 Geopyter.tex
Major rewrite incorporating Reviewer feedback. Still more work to do.

\newcommand{\gp}{\textsc{g}eo\textsc{p}y\textsc{t}e\textsc{r}~\/}
\newcommand{\eg}{e.g.~\/}
\newcommand{\ie}{i.e.~\/}
\newcommand{\comment}[1]{\todo[inline, color=green!40]{#1}}

%% Language and font encodings
\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[hyphens]{url}
\usepackage{fancyvrb} % facny verbatim
\usepackage{textcomp}
\usepackage{upquote}  % Turns quotes in verbatim to straight
\usepackage{varioref} % for vpageref[above]
\usepackage{fullpage}
\usepackage{graphicx} 
\usepackage[font=small,labelfont=bf]{caption}
%% Sets page size and margins
%\usepackage[letter,top=3cm,bottom=2cm,left=3cm,right=3cm,marginparwidth=1.75cm]{geometry}

@@

Although \citet{Donoho2017} traces the origins of data science back to
\citeauthor{Tukey1962}'s \textit{The future of data analysis}
(\citeyear{Tukey1962}), its emergence has led to a flourishing of tools and
programmes as universities seek to engage with this nascent---and
lucrative---interdisciplinary field. One of the most profound practical impacts
of this development has been on the ways in which developers share and execute
code: although systems for combining code, commentary, and results have been
around for some time (\eg R-Markdown) these were intended primarily for
replicating research outputs, not for the online teaching of a new technique or
writing of an interactive introduction to a software library. The `open' ethos
and rapid growth of data science, and the need for a lightweight interactive
code-sharing platform mean that teachers in many disciplines are playing
catch-up on both the methods taught and the formats used.

As a discipline, geography has been acutely aware of these developments: changes
in the volume and extent of geographical data available (\eg
\citealp{Graham2013,gonzalez2013big,reades2016}), and in the number and range of
methods employed to identify patterns in those data (\eg
\citealp{Fan2016,Naik2017,Santibanez2015,Stevens2015,ArribasBel2017}), are
allowing us to tackle questions thought unanswerable just a few years ago. At
the same time, however, there is growing evidence of a `hollowing out' of
traditional domain-specific skills: most of what a trained geographer used to do
with specialist software can now be done using free online resources (\eg
Google's Fusion Tables, MapBox, Carto, ArcOnline), but there remains unmet
demand for graduates who can \textit{code} in order to take advantage of the
latest (geo)algorithms and techniques \citep{Singleton2014,Singleton2016}.

@@

Regardless of whether this is a revolution \citep{Wyly2014,Torrens2010} or an
evolution \citep{Barnes2013,Barnes2014,ArribasBel2018}, there is no question
that teaching students and researchers how to code and how to \textit{think}
computationally presents new challenges
\citep{Etherington2016,Muller2014,rey_09jgs} and that there is little in the way of a tradition of doing so to a high level within Geography (see surveys in \citealp{Bowlick2017,Bowlick2018}). Consequently, the majority of resources used in the
classroom are being developed from scratch at each institution, and often by
newly-appointed lecturers and assistant professors coming from the relatively
small number of elite research facilities active in this domain (see discussion
in \citealp{esrc2013}). Not only has this led to the duplication of effort at
multiple sites, but it has also had an enormous impact on the productivity---as
commonly measured by institutions---of early career researchers, many of whom
are in their first teaching post.

The overarching purpose of this software is therefore to address two seemingly
contradictory issues: the recognition that no two instructors teach in precisely
the same way, and the fact that few instructors have the luxury of both time and
tenure to develop compelling new course material in splendid isolation. This
where the Geographical Python Teaching Resource
(\textsc{g}eo\textsc{p}y\textsc{t}e\textsc{r}) comes into play: although
developed with geographers in mind, it provides a generic means by which
instructors in any discipline can selectively incorporate and re-mix programming
and conceptual content from existing Jupyter notebooks while providing their own
`gloss' for this content. Our approach therefore seeks to develop a richer, open
community of developers and teachers who work \textit{together} on new teaching
materials, and to support instructors in quickly adapting existing materials to
new delivery formats.

@@

Like many open source projects, \gp grew out of a need to scratch an itch: in
our roles as teachers of (geo)computational concepts and methods we grappled on
an annual basis with the demands of developing and updating instructional
material with complex interdependencies. And, while the fundamental computing
and analytic concepts may remain fairly stable, the field is highly-dynamic in
terms of both the content and the rate of evolution of new features in
contributed code upon which we rely for functionality. Currently, every time a
widely-used library package is updated a protracted sequence of revisions
ensues: some teachers, pressed for time, may resist making use of---or even
alerting students to---the new features; others may leap in enthusiastically,
only to discover that they hadn't fully thought through the challenges of
conflicting dependencies and evolving feature sets.

@@

We drew inspiration from the \textsc{foss} iPython project
\citep{perez_2007_ipython} and its transformative impact on the sharing of
integrated code, text, and multimedia resources. And we also looked to the
Python Spatial Analysis Library (\textsc{p}y\textsc{sal}) project
\cite{rey_19sea}in which contributors take ownership of application areas
through managed collaboration instead of a contributory `free-for-all'. The
subsequent emergence of the Jupyter project \cite{kluyver16} from iPython has
led to the rapid growth of a rich ecosystem of tools that support the kinds of
interaction needed to make this an effective teaching resource:
multiuser-notebooks (jupyterhub), automation of grading (nbgrader), interactive
widgets for visualisation (ipywidgets), and packages to facilitate manipulation
of the notebook itself (nbformat and notebook).

With these criteria and resources in mind, we began looking for ways to lower
the `entry costs' for faculty to both take from, and share into, the corpus of
instructional material, as well as for a way to address the
customisation/localisation challenge. \gp was our solution: a single, open hub
of teaching concepts and practical programming tasks that can be flexibly
assembled into classes, or entire courses! In the \gp approach we recognise that
teachers need to be able to develop their own classroom `story' at their own
pace (and in conformance with their institution's teaching pattern), while
supporting them with a library of up-to-date materials from which they can pick
and choose. So in much the same way that a computer application is compiled from
the contributions of many developers, \gp allows individual classes and entire
courses to be `compiled' from the contributions of many skilled teachers and
developers. As such, this positions our work within contextualised programming initiatives \citep{Guzdial2010,Lukkarinen2016} of which the CS+X approach is perhaps best-known \citep[][\eg]{Mir2017}; however, \gp represents a significant advance in this area since it not only builds in context at the level of individual modules, but at the level of a shared set of teaching resources as well!

@@

At least in our experience, the pace of technological and methodological change
is such that the gap between what instructors know and what students know is
substantial, and in many cases might even be growing. The differences within
individual student cohorts may be greater still, however, and while students
need to be able to begin their learning `journey' at some arbitrary point in the
progression from `learning to code' towards `computational social scientist'
\citep{Lazer2009}, instructors require the ability to assemble a course that
meets students where they \emph{are} rather than where the instructor or
institution might \emph{wish} them to be. At the same time, there is also a strong need to support more open-ended exploration during `practicals' \citep{Unwin1980}, particularly by more advanced students who may `tune out' if the means of progression is overly structured and rigid.

@@

Each atom would employ \emph{domain-specific} illustrative examples and code so
as to anchor learning in relevant problems and applications. An atom on arrays
written for geographers could start by showing how these simple data structures
can be used to hold data about cities (\eg name, country, population); but a
subsequent unit could build on this with an an illustration of how nested data
structures allow us to add city location using nested coordinate pairs (\eg
latitude and longitude within an array-of-arrays). This is why managed
collaboration around sets of related concepts is so important: quite aside from
the fact that no \textsc{foss} project succeeds without some level of executive
control, it's critical that \textit{someone} is responsible for ensuring a
degree of consistency in how the units fit together within a set of
closely-related topics.

The purpose of the bottom-up approach is that these units can then be flexibly
assembled into Sessions: from the \textit{same ingredients} (\ie atoms) the
instructor could create two quite different modules---one for delivery in
person, one for delivery online---by organising the elements in different ways.
Quite simply, we don't want to have to rewrite material for each format:
lecture/practical, `flipped' classroom, or \textsc{mooc}. Furthermore, sessions
designed for experts (\eg those pursuing continuing online education) might be
able to `move' students through many more atoms of instruction in a single
session than a similarly laid out (in terms of time) session designed for
first-time programmers in an undergraduate programme. So sessions need to be
able to \emph{incorporate} atoms in a way that minimises the level of effort
involved in finding the `best' way to, for example, explain the concept of
recursion while maximising the ability of the instructor to relate this concept
to the students' practical experience (\eg by providing a `context' that is
anchored in a locally relevant `story' or data).

@@

So we are trying to design a system in which sessions and modules are
\emph{composed} out of atomic units of instruction that can be, optionally,
surrounded by the instructor's own `narrative' or supported by more locally
relevant (geo)data. We therefore want to produce a set of teaching materials
that are highly portable, easily re-used or edited, and that enable the
instructor to select only the elements from which they wish to compose their
materials. As intimated above, \gp operationalises this through the Jupyter
project and its ability to provide in-browser access to the Python
interpreter\footnote{Other languages are also possible using different kernels;
there is no reason that \gp couldn't be used to create instructional materials
in these languages as well.}, and it also takes advantage of the dominance of
Git and GitHub for version control as a means for tracking authorship across
many versions and changes.

@@

Jupyter notebooks are written as a mix of executable code cells and
non-executable text formatted with the widely-used `markdown' syntax. Notebook
structure is provided through headers in markdown cells: a `\#' pre-pended to a
line of text is generally taken to be the title of the notebook; `\#\#' at the
start of a line provides a second level of structure (\ie level 2 headers);
`\#\#\#' indicates level 3 headers; etc. For our purposes, what's relevant is
that these headers yield a natural hierarchy that corresponds closely to the
h1\ldots{}h6 model used by the HTML markup language that lies at the heart of
the World Wide Web. This hierarchy allows us to `abstract out' the problem of
inferring the \emph{meaning} of different sections of the notebook since the
instructor does it for us.

@@

\begin{figure}[hbtp]
  \centering
  \includegraphics[width=\textwidth, angle=0]{Illustration.pdf}
  \caption{Illustration of \gp Notebook `Compilation' Process}
  \label{fig:Illustration}
\end{figure}

@@

The final element of \gp is necessarily a syntax for specifying how Sessions and
Modules should be compiled from Atoms. We've noted the technical mechanism, as
well as the mapping between markdown and \textsc{html} formatting, above but how
do we select some mix of code and markdown material in one notebook to be
incorporated into another? And how do we do this in a way that is both simple to
express and able to resolve ambiguity? Fortunately, such a model already exists
and was hinted at in Figure \ref{fig:Illustration}: Cascading Style Sheets
(\textsc{css}) uses well-understood `selectors' to specify one or more elements
on a web page to which a set of presentational styles should be applied: an `h1'
in a style sheet indicates that all level 1 headers should observe the styling
rules declared immediately afterwards; while `h1.important' specifies that only
a level 1 header of the class `important' should be selected. More powerfully,
`h1.important a.super' selects \emph{only} anchors of the class `super' that
fall `within' (or `under', if you prefer') a level 1 header of the `important'
class.

What is particularly elegant about this approach is that it provides a means for
selecting and styling multiple pieces of content in the document in one
declaration (where this is desirable) \textit{and} a means for disambiguating
content with the same name but in different positions within a document's
hierarchical structure (where it is not). We have essentially repurposed
\textsc{css} as a means of selecting and importing content from one notebook
into another! Unfortunately, the nature of Jupyter notebooks does not allow this
to happen dynamically at run-time, but it does allow something similar to happen
when an instructor is `compiling' new Sessional and Module content. In short,
the instructor writes whatever content they wish but, using syntax similar to
the examples below, wherever they want to incorporate contributed content from
\gp (or elsewhere) they have only to write an `include' statement to select and
import sections of another notebook. This approach also works recursively: a
notebook can include content from a notebook that itself includes content from
another notebook\ldots

Adopting the language used above: we envision an Atom as a short notebook
focussing on a single concept or method (\eg lists, object-oriented design, or
spatial autocorrelation); some or all of each Atom can then be selected and
imported into a Session, which is itself a notebook; and the sessions can then
be selected and managed through a Module, which can be a notebook or a set of
notebooks. This process is initiated by the instructor creating a blank text
cell in a Jupyter notebook and writing an `include' statement. The statement
should be the \textit{only} content in the cell since \gp will be replacing the
cell with an unknown number of whole text and code cells from the referenced
notebook.

Crucially, `include' statements can be freely intermingled with the instructor's
own content, allowing the instructor to `frame' the concepts in a way that suits
their teaching style but which saves them having to reinvent the wheel for each
class. A Session tackling standardisation could include elements of the relevant
Atom from \gp while still allowing the instructor to interject comments,
observations, questions, and additional tasks to ground the learning experience
in the local context (individual, institutional, etc.). To illustrate this more
clearly, an Atom on Python's approach to dealing with lists could be
incorporated into a longer Session as follows:

\begin{Verbatim}[fontsize=\small]
@include {
    'nb'     = 'http://geopyter.org/atoms/fundamentals/lists.ipynb',
    'select' = 'h1.Understanding Lists'
}
\end{Verbatim}

The `nb' is the path---local or remote---to a valid Jupyter notebook from which
the Session or Module developer wants to import content. The `select' parameter
specifies a selector for which the \gp tool will search within the source
notebook. All content from that point onwards \emph{up to the next selector at
the same level} will the then be copied into the compiled notebook. In example
above, if there were a following h1 covering, for example, `List Operators' then
this would \emph{not} be included because, from a structural standpoint, it is
of equal importance (at the same level in the hierarchy) to `Understanding
Lists' but has \emph{not} been selected. Furthermore, any h2 or h3 subsections
within the `Understanding Lists' section \textit{would} be included since they
are presumed to be providing pedagogical and logical structure to the
Understanding Lists section and so should be carried over.

\begin{Verbatim}[fontsize=\small]
@include {
    'resource' = 'http://geopyter.org/atoms/fundamentals/lists.ipynb',
    'select' = 'h1.Understanding Lists -h3.Lists Example; h1.Using Lists -h3.Another Example'
}
\end{Verbatim}

@@

There has been increasing interest at universities in Massive Open Online
Courses, or \textsc{mooc}s, as vehicles for expanding access to higher education
through online delivery. For HE managers, the \textsc{mooc} promises both
enhanced revenue---depending on the model this implies either more people paying
to take the modules or more paying for a certificate of completion at the
end---and enhanced access as people who would be unable to attend the university
in person are nonetheless able to pursue a degree. Our own experience suggests
substantial student interest in `computational social science'
\citep{Lazer2009}, with international students being the most keen on such
modules as they are seen to provide marketable skills (programming) for
graduates from a discipline (geography) with generally high employment rates
(see \citeauthor{rgs2017} \citeyear{rgs2017}).

@@

Because it builds on \textsc{foss} approaches to software development, we expect
\gp to benefit from network effects: the more people use it, the more useful it
becomes, and the more people use it. In common with many such projects we
\emph{do} expect that a relatively small subset of users will contribute the
majority of the content; however, unlike a traditional software project there is
an important role for \emph{teachers}, not just highly-skilled \emph{developers}
and we think that this represents a really exciting opportunity for innovative
new approaches to rise to the surface. There may be only a few who can write the
Python code to conduct a Geographically-Weighted Regression analysis, but it
will be interesting to see how many creative, insightful ways there are to
explain it!

>>>>

\newcommand{\gp}{\textsc{g}eo\textsc{p}y\textsc{t}e\textsc{r}~\/}
\newcommand{\eg}{e.g.~\/}
\newcommand{\ie}{i.e.~\/}
\newcommand{\pysal}{\textsc{p}y\textsc{sal}~\/}
\newcommand{\comment}[1]{\todo[inline, color=green!40]{#1}}

%% Language and font encodings
\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[hyphens]{url}
\usepackage{fancyvrb} % facny verbatim
\usepackage{textcomp}
\usepackage{upquote}  % Turns quotes in verbatim to straight
\usepackage{varioref} % for vpageref[above]
\usepackage{fullpage}
\usepackage{graphicx} 
\usepackage{tabularx}
\newcolumntype{s}{>{\hsize=.25\hsize \raggedright\arraybackslash}X}
\newcolumntype{b}{>{\hsize=.75\hsize}X} 	
\usepackage{booktabs}
\usepackage{wrapfig}
\usepackage[font=small,labelfont=bf]{caption}
%% Sets page size and margins
%\usepackage[letter,top=3cm,bottom=2cm,left=3cm,right=3cm,marginparwidth=1.75cm]{geometry}

@@

Although \citet{Donoho2017} traces the origins of data science back to
\citeauthor{Tukey1962}'s \textit{The future of data analysis}
(\citeyear{Tukey1962}), it is only recently that this set of previously disparate practices for working with large data sets has begun to be formalised in a way that might allow it to be taught in a university context instead of acquired on the job through `learning-by-doing'. One of the most profound practical impacts of this development has been on the ways in which developers share and execute code: although systems for combining code, commentary, and results have been around for some time (\eg R-Markdown) these were intended primarily for replicating research outputs and not for the teaching of a new technique or writing of an interactive introduction to a software library. The emergence of data science---and the overall pace of change in the tools and programming libraries that it employs---gave new impetus to the search for lightweight ways of sharing code and documentation, and an interactive browser-based technology called Jupyter was the result.

As a result, university educators in many disciplines are playing catch-up with both the methods and the tools. Geography has been acutely aware of these developments: changes in the volume and extent of spatial data available (\eg \citealp{Graham2013,gonzalez2013big,reades2016}), and in the number and range of methods employed to identify patterns in those data (\eg \citealp{Fan2016,Naik2017,Santibanez2015,Stevens2015,ArribasBel2017}), are allowing us to tackle questions thought unanswerable just a few years ago. At the same time, there is growing evidence of a `hollowing out' of once domain-specific skills \citep{Singleton2014,Singleton2016}: most of what a trained geographer used to do with specialist Geographic Information Systems (\textsc{gis}) software can now be done using free online resources (\eg Google's Fusion Tables, MapBox, Carto, ArcOnline). However, there remains unmet demand for graduates who can \emph{code} using the latest (geo)algorithms and techniques, so in this article we present a means of bringing the benefits of data science-approaches to instruction using Jupyter `notebooks' to the teaching of Python programming and geospatial analysis: Geographical Python Teaching Resources (GeoPyTer).

@@

Regardless of whether this is a revolution \citep{Wyly2014,Torrens2010} or an evolution \citep{Barnes2013,Barnes2014,ArribasBel2018}, there is no question that teaching students and researchers how to code and how to \textit{think} computationally presents new challenges \citep{Etherington2016,Muller2014,rey_09jgs} and that there is little in the way of a tradition of doing so to a high level within Geography (see surveys in \citealp{Bowlick2017,Bowlick2018}). Consequently, the majority of resources used in the classroom are being developed from scratch at each institution, often by newly-appointed lecturers and assistant professors coming from the relatively small number of elite research facilities active in this domain (see discussion in \citealp{esrc2013}). Not only has this led to the duplication of effort at multiple sites, but it has also had an enormous impact on the productivity of early career researchers, many of whom are in their first teaching post, as commonly measured by universities.

The overarching purpose of this software is therefore to address two seemingly
contradictory issues: the recognition that no two instructors teach in precisely
the same way, and the fact that few instructors have the luxury of both time and
tenure to develop compelling new course material in splendid isolation. This is 
where the Geographical Python Teaching Resource
(\textsc{g}eo\textsc{p}y\textsc{t}e\textsc{r}) comes into play: although
developed with geographers in mind, it provides a generic means by which
instructors in any discipline can selectively incorporate and re-mix programming
and conceptual content from existing Jupyter notebooks while providing their own
`gloss' for this content. Our approach therefore seeks to develop a rich community of developers and teachers who work \textit{together} on new teaching materials, who make their output entirely \emph{open} to the wider community, and whoe support instructors in quickly adapting existing materials to new delivery formats.

@@

Like many open source projects, \gp grew out of a need to scratch an itch: in our roles as teachers of (geo)computational concepts and methods we grappled on an annual basis with the demands of developing and updating instructional material with complex interdependencies. And, while the fundamental computing and analytic concepts may remain fairly stable, the field is highly-dynamic in terms of both the content (\ie what should be in the curriculum) and the code (\ie how applications should be taught). Currently, every time a widely-used library package is updated a protracted sequence of revisions ensues: some teachers, pressed for time, may resist making use of---or even alerting students to---the new features; others may leap in enthusiastically, only to discover that they hadn't fully thought through the challenges of conflicting dependencies and evolving feature sets.

@@

We drew inspiration from the \textsc{foss} iPython project
\citep{perez_2007_ipython} and its transformative impact on the sharing of
integrated code, text, and multimedia resources. And we also looked to the
Python Spatial Analysis Library (\textsc{p}y\textsc{sal}) project
\cite{rey_19sea} in which contributors take ownership of application areas
through managed collaboration instead of a contributory `free-for-all'. The
subsequent emergence of the Jupyter project \citep{kluyver16} from iPython has
led to the rapid growth of a rich ecosystem of tools that support the kinds of
interaction needed to make this an effective teaching resource:
multiuser-notebooks (jupyterhub), automation of grading (nbgrader), interactive
widgets for visualisation (ipywidgets), and packages to facilitate manipulation
of the notebook itself (nbformat and notebook).

With these criteria and resources in mind, we began looking for ways to lower
the `entry costs' for faculty to both take from, and share into, the corpus of
instructional material, as well as for a way to address the
customisation/localisation challenge. \gp was our solution: a single, open hub
of teaching concepts and practical programming tasks that can be flexibly
assembled into classes, or entire courses! In the \gp approach we recognise that
teachers need to be able to develop their own classroom `story' at their own
pace (and in conformance with their institution's teaching pattern), while
supporting them with a library of up-to-date materials from which they can pick
and choose. So in much the same way that a computer application is compiled from
the contributions of many developers, \gp allows individual classes and entire
courses to be `compiled' from the contributions of many skilled teachers and
developers. As such, this positions our work within contextualised programming initiatives \citep{Guzdial2010,Lukkarinen2016} of which the CS+X approach is perhaps best-known \citep[\eg][]{Mir2017}; however, \gp represents a significant advance in this area since it not only builds in context at the level of individual modules, but at the level of a shared set of teaching resources as well!

@@

At least in our experience, the pace of technological and methodological change
is such that the gap between what instructors know and what students know is
substantial, and in many cases might even be growing. The differences within
individual student cohorts may be greater still, however, and while students
need to be able to begin their learning `journey' at some arbitrary point in the
progression from `learning to code' towards environmental, geospatial, or computational social scientist \citep{Lazer2009}. Instructors require the ability to assemble a course that meets students where they \emph{are} rather than where the instructor or institution might \emph{wish} them to be. At the same time, there is also a strong need to support more open-ended exploration during `practicals' \citep{Unwin1980}, particularly by more advanced students who may `tune out' if the means of progression is overly structured and rigid\ldots or just too slow to keep their attention.

@@

\begin{wraptable}{r}{0.6\textwidth}
\centering
\caption{Overview of System Components}
\label{table:1}
\begin{tabularx}{0.6\textwidth} {sb}
 \toprule
 \textbf{Concept} & \textbf{Usage} \\
 \midrule
 Atom  & A `teachable idea' combining explanation and code for a core, typically singular, concept (\eg Variables, Lists, Projections, Robust Rescaling\ldots). \\
 Sessions  & A `teachable unit' combining \textbf{multiple atoms} for delivery in a particular learning context (\eg Beginner Lab-based Course, Intermediate Flipped Module, Online Course\ldots) \\
 Modules  & A \textbf{sequence of sessions} designed to achieve one or more pedagogical objective and/or required for credit in a particular education context. \\
\bottomrule 
\end{tabularx}
\end{wraptable}

Each atom would employ \emph{domain-specific} illustrative examples and code so as to anchor learning in problems and applications relevant to the learner. An Atom on arrays written for geographers could start by showing how these simple data structures can be used to hold data about cities (\eg name, country, population); but a subsequent unit could build on this with an an illustration of how nested data structures allow us to add city location using nested coordinate pairs (\eg latitude and longitude within an array-of-arrays). This is why managed collaboration around sets of related concepts is so important: quite aside from the fact that no \textsc{foss} project succeeds without some level of executive control, it's critical that \textit{someone} is responsible for ensuring a degree of consistency in how the units fit together within a set of closely-related topics.

The purpose of the bottom-up approach is that these units can then be flexibly assembled into Sessions: from the \textit{same ingredients} (\ie Atoms) the instructor could create two quite different modules---one for delivery in person, one for delivery online---by organising the elements in different ways. Quite simply, we don't want to have to rewrite material for each format: lecture/practical, `flipped' classroom, or online course. Furthermore, sessions designed for experts (\eg those pursuing continuing online education) might be able to `move' students through many more atoms of instruction in a single session than a similarly laid out (in terms of time) session designed for first-time programmers in an undergraduate programme. So sessions need to be able to \emph{incorporate} atoms in a way that minimises the level of effort involved in finding the `best' way to, for example, explain the concept of recursion while maximising the ability of the instructor to relate this concept to the students' practical experience (\eg by providing a `context' that is anchored in a locally relevant `story' or data).

@@

So we are trying to design a system in which sessions and modules are
\emph{composed} out of atomic units of instruction that can be, optionally,
surrounded by the instructor's own `narrative' or supported by more locally
relevant (geo)data. We therefore want to produce a set of teaching materials
that are highly portable, easily re-used or edited, and that enable the
instructor to select only the elements from which they wish to compose their
materials. As intimated above, \gp operationalises this through the Jupyter
project and its ability to provide in-browser access to the Python
interpreter\footnote{Other languages are also possible using different kernels;
there is no reason that \gp couldn't be used to create instructional materials
in these languages as well.}, and it also takes advantage of the dominance of
the Git tool---and the GitHub service/web site---for version control as a means for tracking authorship across many versions and changes.

@@

Jupyter notebooks are written as a mix of executable code cells and
non-executable text formatted with the widely-used `markdown' syntax. Notebook
structure is provided through headers in markdown cells: a `\#' pre-pended to a
line of text is generally taken to be the title of the notebook; `\#\#' at the
start of a line provides a second level of structure (\ie level 2 headers);
`\#\#\#' indicates level 3 headers; etc. For our purposes, what's relevant is
that these headers naturally yield a semantic hierarchy that corresponds closely to the h1\ldots{}h6 model used by the HTML markup language that lies at the heart of the World Wide Web. This hierarchy allows us to `abstract out' the problem of inferring the \emph{meaning} of different sections of the notebook since the instructor does it for us through their use of headers.

\subsection{Approach}

In order to compile Atoms into Sessions and Modules, \gp necessarily requires a compositional syntax. We've noted the conceptual mapping between markdown and \textsc{html} formatting, above but how do we select some mix of code and markdown material in one notebook to be incorporated into another? And how do we do this in a way that is both simple to express and able to resolve ambiguity? Fortunately, such a model already exists and was hinted at in Figure \ref{fig:Illustration}: Cascading Style Sheets (\textsc{css}) uses well-understood `selectors' to specify one or more elements on a web page to which a set of presentational styles should be applied.

With \textsc{css} an `h1' in a style sheet indicates that all \textsc{html} Level 1 Headers should observe the styling rules declared immediately afterwards; while `h1.important' specifies that only a Level 1 Header of the class `important' should be selected. In fact, \textsc{css} also allows for nested selectors: for some element(s) of a `larger' selection to be selected in turn. This is normally used to do things like specify mouseover behaviours for a menu: that all anchors (\ie links) that are within a division of class \texttt{menu} should act in \textit{this} way when the mouse passes over them. What is particularly elegant about this approach is that it provides a means for selecting and styling multiple pieces of content in the document in one declaration (where this is desirable) \textit{and} a means for disambiguating content with the same name but in different locations within a document's hierarchical structure (where it is not).

Conceptually, \gp reframes this in terms of the meaning imparted by the instructor to the notebook: all cells coming after a Level 1 Header are considered to be part of that element's semantic field until another Level 1 Header is encountered or the end of the document is encountered, whichever comes first. So a Level 2 Header coming `after' (or `below', if you prefer) a Level 1 Header is still part of that higher-level element's semantic field, \textit{but} we can select it uniquely within the notebook using the standard \textsc{css} form of \texttt{h1.content h2.subcontent}. And with this we have essentially repurposed \textsc{css} as a means of selecting and importing content from one notebook into another! This is illustrated in schematic form in Figure \ref{fig:structure} where the \texttt{>} imply a hierarchical relationship.

\begin{wrapfigure}{r}{0.51\textwidth}
  \centering
  \caption{Illustration of Atom Structure}
  \includegraphics[width=0.5\textwidth]{Single_Document.pdf}
  \label{fig:structure}
\end{wrapfigure}

Unfortunately, the nature of Jupyter notebooks does not allow this to happen dynamically at run-time, but it does allow something similar to happen when an instructor is `compiling' new Sessional and Module content. In short, the instructor writes whatever content they wish but, using syntax similar to the examples below, wherever they want to incorporate contributed content from \gp (or elsewhere) they have only to `include' it by specifying both a source and a selection. And this approach also works recursively: a notebook can include content from a notebook that itself includes content from another notebook\ldots

Adopting the approach outlined above: we typically envision an Atom as a short notebook focussing on a core concept or method (\eg lists, object-oriented design, or spatial autocorrelation); some or all of each Atom can then be selected and imported into a Session, which is itself a notebook; and the sessions can then be selected and managed through a Module, which can be a notebook or a set of notebooks. This process is initiated by the instructor creating a blank text cell in a Jupyter notebook and writing an `include' statement. The statement should be the \textit{only} content in the cell since \gp will be replacing the cell with an unknown number of whole text and code cells from the referenced notebook.

Crucially, \texttt{include} statements can be freely intermingled with the instructor's own content, allowing the instructor to `frame' the concepts in a way that suits their teaching style but which saves them having to reinvent the wheel for each class. A Session tackling standardisation could include elements of the relevant Atom from \gp while still allowing the instructor to interject comments, observations, questions, and additional tasks to ground the learning experience in the local context (individual, institutional, etc.). To illustrate this more clearly, an Atom on Python's approach to dealing with lists could be incorporated into a longer Session as follows:

\begin{Verbatim}[fontsize=\small]
@include {
    'nb'     = 'http://geopyter.org/atoms/fundamentals/lists.ipynb',
    'select' = 'h1.Lists'
}
\end{Verbatim}

\texttt{nb} is a path---local or remote---to a valid Jupyter notebook from which the Session or Module developer wants to import content. The \texttt{select} parameter specifies a selector for which the \gp tool will search within the source notebook. All content from that point onwards \emph{up to the next selector at he same level} will the then be copied into the compiled notebook. In example above, if there were a following \texttt{h1} covering, for example, `List Operators' then this would \emph{not} be included because, from a structural standpoint, it is of equal importance (at the same level in the hierarchy) to `Lists' but has \emph{not} been selected. Furthermore, any \texttt{h2} or \texttt{h3} subsections within the `Lists' section \textit{would} be included since they are presumed to be providing pedagogical and logical structure to the Understanding Lists section and so should be carried over.

\begin{Verbatim}[fontsize=\small]
@include {
    'resource' = 'http://geopyter.org/atoms/fundamentals/lists.ipynb',
    'select' = 'h1.Lists -h3.Lists Example; h1.List Operators -h2.Concatenation'
}
\end{Verbatim}

\subsection{Putting it All Together}

@@

\begin{figure}[hbtp]
  \centering
  \includegraphics[width=\textwidth, angle=0]{Multiple_Documents.pdf}
  \caption{Illustration of \gp Notebook `Compilation' Process}
  \label{fig:Illustration}
\end{figure}

@@

@@

There has been increasing interest at universities in Massive Open Online
Courses, or \textsc{mooc}s, as vehicles for expanding access to higher education
through online delivery. For managers in Higher Education (\textsc{he}), the \textsc{mooc} promises both
enhanced revenue---depending on the model this implies either more people paying
to take the modules or more paying for a certificate of completion at the
end---and enhanced access as people who would be unable to attend the university
in person are nonetheless able to pursue a degree. Our own experience suggests
substantial student interest in `computational social science'
\citep{Lazer2009}, with international students being the most keen on such
modules as they are seen to provide marketable skills (programming) for
graduates from a discipline (geography) with generally high employment rates
(see \citeauthor{rgs2017} \citeyear{rgs2017}).

@@

Because it builds on \textsc{foss} approaches to software development, we expect \gp to benefit from network effects: the more people use it, the more useful it becomes, and the more people use it. The open source, peer-generated approach would, however, also be likely to present something of a challenge over time: as more people seek both to use and to contribute to \gp we would be likely to see the emergence of divergent norms, examples, and data across Atoms, Sessions, and Modules. In other words, the coherence of \gp as a teaching platform would become a victim of any real success in terms of widespread adoption.

\subsection{Coordination}

Here we believe that the \pysal project offers a useful template. Although \pysal is a free open source project, it is \emph{not} a free-for-all: domain specialists tend to gravitate towards those parts of the project to which they have the most to contribute and, over time, those who coordinate and enable the most substantive contributions to the codebase in terms of features and performance tend to become the `managers' of individual components of the overall tool (\eg lib, model, explore, viz). Overall project coherence is maintained via regular calls and online discussion boards, and a synchronised release schedule so that changes can be coordinated and knock-on effects identified.

We expect \gp to follow a similar path in that a relatively small subset of users will contribute the majority of the content; however, unlike a traditional software project there is an important role for \emph{teachers}, not just highly-skilled \emph{developers}. We think that this represents a really exciting opportunity for innovative new approaches to rise to the surface. There may be only a few who can write the Python code to conduct a Geographically-Weighted Regression analysis, but it will be interesting to see how many creative, insightful ways there are to explain it!

The educational focus implies that it may well be the most committed teachers who end up coordinating and reviewing contributions to \gp from other instructors and developers. We would also argue for the relevance of grouping sets of Atoms together in basic libraries such as `foundations' (the basics of variables and data structures); `description' (describing data); etc. Within these sets, contributions would be managed by a coordinator to ensure that examples, style, and other features are consistent for use in a set of Sessions introducing Python programming, or Local Indicators of Spatial Autocorrelation.

The key to \gp is that it seeks to balance the benefits of code-sharing with those of local expertise: the instructor is free to write their own exegesis, if you will, of the code and its relevance to a particular Session or Module, but the burden of developing a cogent, domain-specific demonstration can be shared with others and compelling examples more widely adopted without the effort of reinventing the wheel. Where irreconcilable differences arise between pedagogical approaches or models, then we might expect to see small groups of collaborators `fork' the codebase and offer their own models; this is, of course, a valid approach with open source code and one from which all instructors can ultimately benefit!

\subsection{Credit}
